The Noisy Work of Uncertainty Visualisation Research: A Review

### 2024-07-24 paper.qmd
edited "why is uncertainty visualisation difficult" section

This consensus leads us to understand that uncertainty visualisation is motivated by the need for a sort of "visual hypothesis test". A successful uncertainty visualisation would act as a "statistical hedge" for any inference me make using the graphic. This implies three key criteria for a good uncertainty visualisation:

  1) Reinforce justified conclusions to encourage confidence in results
  2) Prevent unjustified conclusions that are indistinguishable from noise
  3) Perform tasks 1) and 2) in a way that is proportional to the level of confidence in those conclusions.

We can refer to the we draw from a graphic to be its "signal" and the variance that makes this signal harder to identify as the "noise". As @fig-choropleth showed, visualisations that are unconcerned with uncertainty have no issue showing justified signals, but struggle with the display of unjustified signals. Therefore, we coin this approach to uncertainty visualisation as "signal-supression" since it primarily differentiates itself from from the "noiseless" visualisation approach through criteria (2). That is, the main difference between an uncertainty visualisation and a "normal" visualisation, is that an uncertainty visualisation should prevent us from drawing unjustified conclusions.

@@

@fig-bivariate a variation of the choropleth map, where we have a two dimensional colour palette. Not only is temperature mapped to hue, but variance is also mapped to saturation. While these two maps *do* look visually different (which was not the case in the choropleth map) the spatial trend is still clearly visible in both graphics. This means the uncertainty *is* being communicated, however the primary take away in the graphic is the spatial trend (that does not exist). The graphic no longer has a transparency issue as the uncertainty *is* technically included, but is this a uncertainty visualisation? And if not, why?

@@

## What makes uncertainty visualisation difficult?
Uncertainty visualisation is a uniquely difficult problem in visualisation. @Hullman2016 commented that it is straightforward to show a value but it is much more complex to show uncertainty. There are two primary theories as to *why* uncertainty visualisation is unique.

The first group believes uncertainty visualisation are unique due to the psychological heuristics involved in interpreting uncertainty [@Spiegelhalter2017; @Hullman2019]. While it is true that heuristics lead people to avoid uncertainty [@Spiegelhalter2017] there is no evidence that this psychological effect translates to a problem in visualisation. Were this the case, visualisations of death, poverty, and any other variable that elicits poor feelings would require its own field of visualisation.

The second group sees uncertainty as a high dimensional visualisation problem as the difficulty is in working out how to add uncertainty into already existing graphics [@moritz2017trust; @Griethe2006]. If we view uncertainty visualisation as a high dimension problem we include uncertainty as we would any other variable, and allow for the signal to be interpreted independent of the variance. This prevents uncertainty and signal being read together as a "single integrated uncertain value" [@Kinkeldey2014] and allows for uncertainty to be ignored [@uncertchap2022]. Additionally, even when reading a map that has a more complicated colour palette than a choropleth map (such as a bivariate map) participants still read the map according to the same principles of the choropleth map [@Ndlovu2023]. This means that even when uncertainty is technically included in the graphic, the take away can still be invalid if the map is read incorrectly. The signal-suppression needs to be communicated through the "gist" of the graphic, as that is primary purpose of a visualisation [@Spiegelhalter2017].  The @fig-bivariate plot's "gist" is "this data has a spatial trend and the estimates have a large variance". It is clear the uncertainty and signal are being communicated separately rather that as a single uncertain value.

This highlights the key problem of viewing uncertainty visualisation as a simple high dimensional problem. While we are trying to include a new variable, uncertainty cannot be treated the same as any other variable. Typically we want visual channels to be separable, that is, the data represented in one visual channel does not interfere with the others, and most research in this area is for this goal [@Smart2019]. Uncertainty visualisation, however, should endeavor for integrability, that is, the channels interfere with one another to allow the uncertainty to suppress the signal. While most aesthetics *are* read separately, colour hue and brightness seem to be integrable [@Vanderplas2020] so even though they can be individually manipulated, they are processed by the brain as a single visual channel. Additionally color value has a more intuitive connection to uncertainty than hue [@Maceachren2012], but it should be mapped such that lighter colours are more uncertain than darker colours. If we could utilise integrability effectively, them the two dimensional space defined by colour value and colour hue could be mentally "flattened" into a single dimension of "statistically valid signal".

@fig-bivariate2 shows a bivariate colour map where uncertainty is mapped to colour value, and temperature is mapped to colour hue. It is clear the spatial trend is still visible in the high uncertainty case. This does not necessarily mean integrability will never be a useful tool for signal suppression, but without a strong understanding of how these visual channels collapse down into a single channel, it may not be a reliable option. Additionally, adding so many variations to colour in a map might make the graphics harder to read and suppress statistically significant signals, while also hurting the plots accessibility [@Vanderplas2015]. A more viable option might be to warp the space uncertainty and signal are mapped to ourselves.

@@

>>>>

This consensus leads us to understand that uncertainty visualisation is motivated by the need for a sort of "visual hypothesis test". A successful uncertainty visualisation would act as a "statistical hedge" for any inference me make using the graphic. Additionally, since the purpose of a visualisation is to give a quick "gist" of the information [@Spiegelhalter2017], this hedging needs to be communicated visually without any need for computation from the viewer. Additionally, @Ndlovu2023 found that participants applied the same methods they used for simple choropleth maps to complicated uncertainty maps even if that take away was invalid. Therefore, this hedging effect needs to be communicated simply through the visualisation. If we refer to the conclusion we draw from a graphic to be its "signal" and the variance that makes this signal harder to identify as the "noise", we can summaries this information into three key requirements. A good uncertainty visualisation needs to:

  1) Reinforce justified signals to encourage confidence in results
  2) Hide signals that are just noise to prevent unjustified conclusions
  3) Perform tasks 1) and 2) in a way that is proportional to the level of confidence in those conclusions.

As @fig-choropleth showed, visualisations that are unconcerned with uncertainty have no issue showing justified signals, but struggle with the display of unjustified signals. Therefore, we coin this approach to uncertainty visualisation as "signal-suppression" since it primarily differentiates itself from from the "noiseless" visualisation approach through criteria (2). That is, the main difference between an uncertainty visualisation and a "normal" visualisation is that an uncertainty visualisation should prevent us from drawing unjustified conclusions.

@@

@fig-bivariate a variation of the choropleth map, where we have a two dimensional colour palette. Not only is temperature mapped to hue, but variance is also mapped to saturation. While these two maps *do* look visually different (which was not the case in the choropleth map) the spatial trend is still clearly visible in both graphics. This means the uncertainty *is* being communicated, however the primary take away in the graphic is the spatial trend (that does not exist). The graphic no longer has a transparency issue as the uncertainty *is* technically included, but is this a good uncertainty visualisation? And if not, why?

@@

## What makes uncertainty visualisation difficult?
Uncertainty visualisation is a uniquely difficult problem. @Hullman2016 commented that it is straightforward to show a value but it is much more complex to show uncertainty. There does not seem to be a consensus on *why* this is the case, but there are two main groups of thought that emerge when trying to justify the difficulty in uncertainty visualisation.

The first group believes uncertainty visualisation are unique due to the psychological heuristics involved in interpreting uncertainty [@Spiegelhalter2017; @Hullman2019]. While it is true that heuristics lead people to avoid uncertainty [@Spiegelhalter2017] there is no evidence that this psychological effect translates to a problem in visualisation. Were this the case, visualisations of death, poverty, and any other variable that elicits poor feelings would require their own field of visualisation.

The second group sees uncertainty as a high dimensional visualisation problem as the difficulty is in working out how to add uncertainty into already existing graphics [@moritz2017trust; @Griethe2006]. If we view uncertainty visualisation as a high dimension problem we include uncertainty as we would any other variable and use high-dimension visualisation techniques to incorporate it. This is the approach taken in @fig-bivariate.

The problem with the high dimension approach is that it treats uncertainty the same as we would any other variable. Typically we want visual channels to be separable, that is, the data represented in one visual channel does not interfere with the others [@Smart2019]. Unfortunately, this means that applying normal visualisation methods to "uncertainty visualisation" will prevent uncertainty and signal being read together as a "single integrated uncertain value" [@Kinkeldey2014], which is the goal of signal-suppression. Separating the uncertainty and the signal also allows for the uncertainty to be ignored, which is a common problem faced by most uncertainty visualisations [@uncertchap2022]. We can see this problem in @fig-bivariate, as it sends the message "this data has a spatial trend and the estimates have a large variance" rather than hiding the statistically invalid signal.

If thses high-dimensional approaches were to work, the uncertainty and the estimates would need to be mapped to integrable channels. That is, the channel that uncertainty and the estimates are mapped to would need to be separately manipulable, but read as a single channel by the human brain. While most aesthetics *are* seperable, there are some variables that have been shown to be integrable, such as colour hue and brightness [@Vanderplas2020]. We can more effectively approach this method by also leveraging visual semiotics and mapping more uncertain values to lighter colors [@Maceachren2012].

@@

@fig-bivariate2 shows a bivariate color map where uncertainty is mapped to color value, and temperature is mapped to color hue. It is clear the spatial trend is still visible in the high uncertainty case. This does not necessarily mean integrability will never be a useful tool for signal suppression, but without a strong understanding of how these visual channels collapse down into a single channel, it may not be a reliable option. Relying on integrability could create unintended consequences of uncertainty visualization, such as displaying phantom signals or hiding justified signals. Additionally, multi-dimensional colour palettes can make the graphics harder to read and hurt the accessibility of the plots [@Vanderplas2015]. A more viable option might be to warp the space uncertainty and signal are mapped to ourselves.
